---
title: |
  | Difference-in-Difference-in-Differences
  | CFPLs Case Study 
pagetitle: DDD CFPLs
subtitle: |
  [Pdf version of this document](https://github.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs/blob/master/CFPLs_DDD.pdf)
  
  [Github repository for this document](https://github.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs)


author: "Chandler Lutz"
date: "Last Updated `r Sys.Date()`"
output:
  pdf_document:
    toc: false     ##No table of contents for pdf documents

  html_document:
    toc: true      ##Use a table of contents for html documents
    toc_depth: 2
    number_sections: true
urlcolor: blue

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- css style for header fonts -->
<style type = "text/css">
h1 { /* Header 1 */
  font-size: 24px;
  color: Black;
}
h2 { /* Header 2 */
    font-size: 18px;
  color: Black;
}
</style>

# Overview

This case study will use a difference-in-difference-in-differences
(DDD) research design to study the impact of the 2008 California
Foreclosure Prevention Laws (CFPLs) that aimed to reduce foreclosures
at the height of the financial crises. These laws are called the
California Foreclosure Prevention Laws. We first start with a classic
difference-in-differences analysis and then cover the DDD approach.

# Preliminaries

* This tutorial is based on [Gabriel et al. (2017)](https://chandlerlutz.github.io/publication/california-foreclosure-prevention-laws/)
* [The Github Repository for this project is here](https://github.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs).
*
  [Download the most recent pdf version here](https://github.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs/blob/master/CFPLs_DDD.pdf).
*
  [Data can be downloaded in either rds or csv here](https://github.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs/tree/master/data-raw).
  
This tutorial will employ the `R` statistical computing environment
and the `data.table` and `magrittr` packages for data manipulation as
well as the `AER` package for regression analysis.

# DDD Notes and resources 

* See this
  [CrossValidated Question](https://stats.stackexchange.com/a/125266/12053) for
  an overview of the usual difference-in-differences methodology 
*
  [Wooldridge Lecture Notes](http://www.eief.it/files/2011/10/slides_6_diffindiffs.pdf)
*
  [These Lecture Notes](http://econ.lse.ac.uk/staff/spischke/ec524/evaluation3.pdf) (More
  Advanced) 
  
# Policy Background 

In 2008, California housing markets were spiraling
downwards. California Policymakers implemented a set of new
foreclosure laws that increased the pecuniary and time costs of
foreclosure in an attempt to mitigate the rise in foreclosures. Our
aim is to analyze the effects of these policies on foreclosures
(See
[Gabriel et. al. (2017)](https://chandlerlutz.github.io/publication/california-foreclosure-prevention-laws/) for
more details).

# Data 

```{r warning=FALSE,message=FALSE}
#Load packages
library(ggplot2); library(magrittr); library(data.table); library(AER)

#Read in the Data
DT <- fread("https://raw.githubusercontent.com/ChandlerLutz/difference-in-difference-in-differences-CFPLs/master/data-raw/county_forc_cfpl_data.csv")
```

The data contain the following variables:

* `CA` -- an indicator equal to 1 for California and zero otherwise 
* `sand.state` -- an indicator equal to 1 for Sand States (AZ, CA, FL,
  NV), states that experienced the largest boom and bust during the
  2000s
* `state.fips` -- the two-digit state fips code 
* `fips.code` -- the five-digit county fips code 
* `CFPL` -- and indicator that equals `1` for the CFPL treatment
  period 
* `zillow.forc` -- the real estate owned (REO) foreclosures per 10,000
  homes 
* `forc.high` -- and indicator equal to 1 for "high" foreclosure
  counties. Classification of counties as high or low foreclosure
  counties is from [Gabriel et al. (2017)](https://chandlerlutz.github.io/publication/california-foreclosure-prevention-laws/). 
* `hh2000` -- the number of housing units in 2000 from the US Census 


```{r }
head(DT)
```

# Research Goal

Our aim is to estimate the effects of the CFPLs on the foreclosures,
`zillow.forc`. 

# Difference-in-differences (DD)

We can first analyze the effects of the program using a simple DD
setup. The idea behind a DD research design, is that we can estimate
the effects of the policy by comparing the change in mean foreclosures
for counties in California relative to the change in mean foreclosures
for Arizona and Nevada. Specifically, we will use counties in
Arizona and Nevada as the control group and counties in California as
the treatment group. Together, California, Arizona, and Nevada make up
the "Sand States" (note: Florida is also considered a Sand State, but
Florida is not in our dataset). 

Let `CFPL = 0` for the pre-CFPL period and `CFPL = 1` for the CFPL
treatment period. Also, `CA = 1` for California counties and `CA = 0`
for counties in other Sand States. Our DD means estimator is

$$
\hat{\delta} = (\bar{Y}_{\text{CA = 1, CFPL = 1}} - \bar{Y}_{\text{CA =
1, CFPL = 0}}) - (\bar{Y}_{\text{CA = 0, CFPL = 1}} - \bar{Y}_{\text{CA =
0, CFPL = 0}})
$$

$\hat{\delta}$ is the difference-in-differences estimator (the
difference in two differences). 

The first difference is the mean change in $Y$ for California counties
(the treatment group), $(\bar{Y}_{\text{CA = 1, CFPL = 1}} -
\bar{Y}_{\text{CA = 1, CFPL = 0}})$. This first difference is also
often called the "before-after" estimator as we compare $\bar{Y}$ in
California before and after the policy 

The second difference is the mean change in $Y$ for non-California
counties (the control group). 

Taking the difference of the above differences yields the
difference-in-differences estimate 

# Difference-in-differences (DD) means estimate

Let's first look at the summary statistics in California versus the
rest of the sand states, where the other sand states (Arizona and
Nevada) are the control group. This will serve as the basis for our DD
means estimator 

```{r }
dd.means <- DT %>%
    #Get the means by the CA and CFPL dummies for the sand states
    .[sand.state == 1,
      .(mean.zillow.forc = weighted.mean(zillow.forc, w = hh2000)),
      by = .(CA, CFPL)]
print(dd.means)
```

Notice here that we restrict the sample to only include the Sand
States as California is a Sand State and the other Sand States
(Arizona and Nevada) represent our control group. We also use the
weighted mean (weighting by the number of households in 2000) as
counties have notably different populations. 

For California counties during the pre-treatment period (`CA = 0` and
`CFPL = 0`), average foreclosures were `222.7555`. This is similar
to the pre-treatment non-CA mean foreclosure estimate: `206.1253`. 

During the CFPL treatment period (`CFPL = 1`), however, the mean
foreclosure estimate was substantially lower for California counties
versus non-California counties (`895.6155` when `CFPL = 0 & CA = 1`
vs. `1539.1376` when `CFPL = 1 & CA = 0`). After the implementation of
the CFPLs, California mean foreclosures increased to `895.6155` (`CFPL
= 1 & CA = 1`), but the non-California mean foreclosures increased to
`1539.1376` (`CFPL = 1 & CA = 0`). Even though foreclosures increased
in California, they increased by substantially less than compared to
the other Sand States. This is the key to the DD estimator: We compare
the *change* in the treatment group to the *change* in the control
group, hence a "difference-in-differences".

We can plot this data using `ggplot2`. Note that we turn `CA` to a
factor so that `ggplot2` classifies `CA` as two different groups:

```{r }
ggplot(data = dd.means, aes(x = CFPL, y = mean.zillow.forc)) +
    geom_line(aes(color = as.factor(CA))) +
    scale_x_continuous(breaks = c(0, 1))
```

From the graph, we see the following points that summarize our DD
analysis thus far:

1. In the pre-CFPL period (`CFPL = 0`), average foreclosures in both
   the treatment group (`CA = 1`) and control group (`CA = 0`) are
   similar. Thus, the treatment and control groups are similar during
   the pre-treatment period. In the jargon of DD studies, we say that
   "the parallel pre-trends assumption is satisfied". The parallel
   pre-trends assumption, the key assumption for DD studies, says that
   during the pre-treatment period (in our case here during the
   pre-CFPL period), that the treatment and control groups only differ
   by a constant, meaning that their pre-treatment trends are
   "parallel". 
2. Foreclosures increased *both* in California and the other Sand
   States increase following the implementation of the CFPLs. Thus, if
   we just looked at California, it would appear as if the CFPLs were
   ineffective at lowering foreclosures
3. The increase in mean California foreclosures was much smaller than
   the increase in mean foreclosures for the other states. Thus, if we
   use the other Sand States as a counterfactual (the path for California
   in the absence of the policy), the plots shows that the CFPLs
   noticeably reduced foreclosures in California. 
   
Using the above formula for the means DD estimator, we have:

$$
\hat{\delta} = (895.6155 - 222.7555) - (1539.1376 - 206.1253) = -660.1523
$$

$\hat{\delta} = -660.1523$ is our DD and means that there were
660.1523 fewer foreclosures in California due to the CFPLs. We discuss
statistical significance below

# Difference-in-differences (DD) means estimate using regression 

We can also obtain our DD estimate and assess our parallel pre-trends
assumption through a regression. The advantage of the regression is
that it simultaneously outputs the estimates of interest as well as
their standard errors (for statistical significance). Note that we'll
only use data from the Sand States (which contain both CA, or
treatment, and our controls, AZ and NV) and use White standard errors
to correct for any potential heteroskedasticity in our data.

```{r }
lm(zillow.forc ~ CA * CFPL, data = DT[sand.state == 1], weights = hh2000) %>%
    coeftest(vcov = sandwich)
```

The regression output shows us all of the requisite DD output. As
`CA`, `CFPL`, and `CA:CFPL` are all indicator variables, we can
easily connect the output from the regression to our DD means
estimates above: 

* $\bar{Y}_{\text{CA = 0, CFPL = 0}} =$ `(Intercept)` $=$ `206.125` 
* $\bar{Y}_{\text{CA = 1, CFPL = 0}} =$ `(Intercept)` $+$ `CA` $=$
  `206.125` $+$ `16.630` = `222.755`
* $\bar{Y}_{\text{CA = 0, CFPL = 1}} =$ `(Intercept)` $+$ `CFPL` $=$
  `206.125` + `1333.012` = `1539.137`
* $\bar{Y}_{\text{CA = 1, CFPL = 1}} =$ `(Intercept)` $+$ `CA` $+$
  `CFPL` $+$ `CA:CFPL` $=$ `206.125` $+$  `16.630` $+$ `1333.012`
  $+$ `-660.152` = `1539.137` = `895.615`
  
As an exercise, check that this output matches the output from
`dd.means`. 
  
Here's what else we learn from the above regression output:

1. The parallel pre-trends assumption is satisfied as there is not a
   statistically significant difference in pre-treatment foreclosures
   between California and the other Sand States (the coefficient on
   `CA` is small and insignificant.
2. The DD coefficient, `CA:CFPL`, the interaction of the `CA` and
   `CFPL` indicators is large in magnitude and statistically
   significant. Thus there is a statistically significant difference
   in foreclosures between California counties and the controls during
   the CFPL period
3. `-660.152` is the DD estimate and it is statistically significant
   with a t-statistic of `-3.61` 

# Other Difference-in-differences estimators

We can leverage the `forc.high` variable and create other DD
estimators. 

First, we can look at "high" foreclosure counties in California versus
"low" foreclosure counties. The advantage of this estimator is that it
will account for California macro-level trends (e.g. an economic shock
that affects the whole state). 

Here is the regression output. Note that we only use California
counties. 

```{r }
lm(zillow.forc ~ CFPL * forc.high, data = DT[CA == 1], weights = hh2000) %>%
    coeftest(sandwich)
```

The coefficient on `forc.high` is significant, indicating that during
the pre-CFPL period, that high foreclosure counties experienced more
foreclosures than low foreclosure counties (2.73 times more
foreclosures, how did I get that number?). This makes sense, but
suggests that our parallel pre-trends assumption is violated.

Further, the coefficient on `CFPL:forc.high` is positive and
significant, suggesting foreclosures in high foreclosure counties
increased more than in low foreclosure counties. It's hard to know
what this means in terms of the CFPL efficacy as parallel pre-trends
assumption is violated. 

We can construct a third DD estimate by comparing high foreclosure
counties in California to high foreclosure counties in other
states. Note here that we subset the data to `forc.high == 1`. This
yields all "high" foreclosure states across the country 

```{r }
lm(zillow.forc ~ CA * CFPL, data = DT[forc.high == 1], weights = hh2000) %>%
    coeftest(sandwich)
```

The output from this regression is rather encouraging: The parallel
pre-trends assumption is satisfied (the coefficient on `CA` is small
and insignificant), and the DD estimate is negative and
significant. The DD estimate tells us that foreclosures were 570.848
lower per 10,000 homes in high foreclosure California counties,
relative to high foreclosure counties in other states. 

# Difference-in-difference-in-differences (DDD)

The above DD estimates are appealing, but none account for within
California macro-level trends while satisfying the parallel pre-trends
assumption. Here, we're going to explore a comprehensive DDD
estimate. 

Essentially, the DDD estimator is going to (1) compare the change in
means between high and low foreclosure counties within California
(first difference); (2) compare the change in means between high and
low foreclosure counties outside of California (second difference);
and (3) take the difference of the first two differences. In essence,
the DDD estimator is the difference in two difference-in-differences
estimators; hence the name "Difference-in-difference-in-differences"


To see how the DDD estimator works, let's calculate the three
differences separately:

**First Difference**:

The first difference is the *within* California difference between
high a low foreclosure counties: 

```{r }
lm(zillow.forc ~ CFPL * forc.high, data = DT[CA == 1], weights = hh2000) %>%
    coeftest(sandwich)
```

Notice that this was one of our DD estimators from above. The within
California DD estimate is `272.642`.


**Second Difference**:

The second difference is difference between high and low foreclosure
counties for counties *outside* California


```{r }
lm(zillow.forc ~ CFPL * forc.high, data = DT[CA == 0], weights = hh2000) %>%
    coeftest(sandwich)
```

The second difference is thus `1029.0368`

**Third Difference (DDD estimate)**:

Our third difference is the difference in our first two differences:
`272.642272.642` $-$ `1029.0368` $=$ `-756.3948`. 

This is our DDD estimate and yields the estimate of the CFPLs on high
foreclosure counties in California, netting out changes in low
foreclosure California counties and change in high foreclosure
counties in other states relative to low foreclosure counties in other
states. 

**The DDD Regression**:

We can confirm our estimate using a full DDD regression. Note here
that we use the full dataset and that `CA` $\times$ `forc.high` is the
treated group and all other counties (even low counties within
California) are controls. The assumption behind identification here is
that CFPL foreclosure prevention policies should have an outsized
impact on high foreclosure counties. 

```{r }
#We're going to save the model b/c we'll need it later 
ddd.mod <- lm(zillow.forc ~ CFPL * forc.high * CA, data = DT, weights = hh2000)
ddd.mod %>% coeftest(sandwich)
```

A couple of things to note about the DDD regression:

1. We can assess the parallel pre-trends assumption: During the
   pre-treatment period, the mean of foreclosures for non-California,
   high foreclosure counties is `(Intercept) + forc.high`; while the
   pre-treatment mean of foreclosures for high foreclosure California
   counties is `(Intercept) + forc.high + CA + forc.high:CA`. The
   difference, and what we need to test for the parallel pre-trends,
   is $H_0$: `CA + forc.high:CA = 0`. The F-statistics for null that
   `CA + forc.high:CA = 0` is insignificant:

```{r }
linearHypothesis(ddd.mod, "CA + forc.high:CA = 0", vcov = sandwich)
```

2. Also notice in the above regression that the coefficient on `CA` is
   insignificant. Thus, low foreclosure California counties are not
   different from low foreclosure counties in other states during the
   pre-CFPL period.
3. The DDD estimate is `-756.3945` and statistically significant at
   the 1 percent level. This estimate means that foreclosures per
   10,000 homes in high foreclosure, California counties fell by
   -756.3945 due to the CFPLs
4. Generally, in both DD and DDD research designs, you can add extra
   control variables to your regression if "randomization is based on
   covariates" (e.g. the parallel pre-trends hypothesis is satisfied
   after controlling for certain variables) or to lower standard
   errors (recall that if a covariate predicts the dependent variable
   and is included in the regression, then the variance of the
   residuals will be lower and the standard errors will fall)
5. See
   [Gabriel et al. (2017)](https://chandlerlutz.github.io/publication/california-foreclosure-prevention-laws/) for
   a panel data and  time-varying setup of the DDD approach. 
